---
title: "Airport Delays Data Analysis"
output: html_notebook
---

The data is from the Bureau of Transportation Statistics: <a href="https://github.com/priankr/airline-delays-data-analysis/blob/main/airlines.csv">Original Dataset (CSV)</a>

This dataset is all about flights in the united states, including information about the number, length, and type of delays between 2003-2016. The data is reported for individual months at every major airport for every carrier.

```{r}
library(tidyverse)
library(ggplot2)
library(gghighlight)

#Importing the csv file to a books Data Frame
delays <- read.csv("airlines.csv")

#Viewing the Data Frame
head(delays)
```
### Missing Values
```{r}
#Identifying total number of missing values
sum(is.na(delays))
```
There are no missing values in the data.
```{r}
#Identifying the column names
colnames(delays)
```
As there are quite a few columns it might be helpful to look at a some some subsets of the data.

We can look at three subsets specifically:

- Number of Flights Delayed Overall
- Number of Flights Delayed by Reason
- Duration of Flight Delay

```{r}
#Creating a character vector named flights in which we are storing column names
flights <- c("Airport.Code","Airport.Name","Time.Month","Time.Month.Name","Time.Year","Statistics.Flights.Cancelled","Statistics.Flights.Delayed","Statistics.Flights.Diverted","Statistics.Flights.On.Time","Statistics.Flights.Total")

#Creating a data frame with information on number of delays
delays_flights <-delays[flights]

#Creating a character vector named flights in which we are storing column names
reason <- c("Airport.Code","Airport.Name","Time.Month","Time.Month.Name","Time.Year", "Statistics...of.Delays.Carrier","Statistics...of.Delays.Late.Aircraft", "Statistics...of.Delays.National.Aviation.System","Statistics...of.Delays.Security","Statistics...of.Delays.Weather")

#Creating a data frame with information on number of delays due to different reasons
delays_reason <-delays[reason]

#Creating a character vector named duration in which we are storing column names
duration <- c("Airport.Code","Airport.Name","Time.Month","Time.Month.Name","Time.Year","Statistics.Minutes.Delayed.Carrier","Statistics.Minutes.Delayed.Late.Aircraft","Statistics.Minutes.Delayed.National.Aviation.System","Statistics.Minutes.Delayed.Security","Statistics.Minutes.Delayed.Total","Statistics.Minutes.Delayed.Weather")

#Creating a data frame with information on duration of delays
delays_duration <-delays[duration]

```

# Flight Data: Number of Flights Delayed Overall
## Average Flights per Month 
The average percentage of flights cancelled, delayed, diverted and on time between 2003-2016.
```{r}
#Average number of delayed flights per month since 2003
avg_flights_cancelled <- sum(delays_flights$Statistics.Flights.Cancelled)

#Average number of delayed flights per month since 2003
avg_flights_delayed <- sum(delays_flights$Statistics.Flights.Delayed)

#Average number of diverted flights per month since 2003
avg_flights_diverted <- sum(delays_flights$Statistics.Flights.Diverted)

#Average number of flights on time per month since 2003
avg_flights_on_time <- sum(delays_flights$Statistics.Flights.On.Time)

#Creating a Pie Chart 
flights_values = c(avg_flights_cancelled,avg_flights_delayed,avg_flights_diverted,avg_flights_on_time )
flights_labels = c("Cancelled","Delayed","Diverted","On Time")

#Calculating the percent each will be of the total
piepercent<- round(100*flights_values/sum(flights_values), 1)

# Plot the chart.
pie(flights_values, labels = piepercent, main = "Distribution of Flights per Month between 2003-2016",col = rainbow(length(flights_values)))
legend("topright", flights_labels, cex = 0.8,fill = rainbow(length(flights_values)))

```
On average, <b>77.8%</b> of flights were On Time each month and <b>20.2%</b> of flights were Delayed.

## Average Flights per Month by Month of Year
The average percentage of flights cancelled, delayed, diverted and on time per month between 2003-2016.
```{r}
 
#Grouping by month
month <- delays_flights %>% group_by(Time.Month)

#Creating a data frame to store the summarized values of flights cancelled each year
delays_flights_month <- month %>% summarise(Statistics.Flights.Cancelled = mean(Statistics.Flights.Cancelled))

#Renaming the columns
colnames(delays_flights_month)[which(names(delays_flights_month) == "Time.Month")] <- "Month"
colnames(delays_flights_month)[which(names(delays_flights_month) == "Statistics.Flights.Cancelled")] <- "Cancelled"

#Summarizing by number of flights delayed each year
month_delayed <- month %>% summarise(Statistics.Flights.Delayed = mean(Statistics.Flights.Delayed))

#Adding the column to delays_flights_month
delays_flights_month$Delayed <-month_delayed$Statistics.Flights.Delayed

#Summarizing by number of flights delayed
month_delayed <- month %>% summarise(Statistics.Flights.Delayed = mean(Statistics.Flights.Delayed))
delays_flights_month$Delayed <-month_delayed$Statistics.Flights.Delayed

#Summarizing by number of flights diverted
month_diverted <- month %>% summarise(Statistics.Flights.Diverted = mean(Statistics.Flights.Diverted))
delays_flights_month$Diverted <-month_diverted$Statistics.Flights.Diverted

#Summarizing by number of flights on time
month_on_time <- month %>% summarise(Statistics.Flights.On.Time = mean(Statistics.Flights.On.Time))
delays_flights_month$On_Time <-month_on_time$Statistics.Flights.On.Time

#Creating a column to display percentage of flights that were on time, on average
delays_flights_month$Percent_On_Time <-100*delays_flights_month$On_Time/(delays_flights_month$On_Time+delays_flights_month$Diverted+delays_flights_month$Delayed+delays_flights_month$Cancelled)

#Display averages for numbers of flights cancelled, delayed, diverted and on time
round(delays_flights_month) 
```
```{r}
#Identifying month with the greatest percent of flights on time, on overage
round(delays_flights_month[which.max(delays_flights_month$Percent_On_Time),])
```
```{r}
#Identifying year with the least percent of flights on time, on overage
round(delays_flights_month[which.min(delays_flights_month$Percent_On_Time),])
```
On average, flights were on time each month the most in <b>September</b> and they were on time the least in <b>December.</b>

## Average Flights per Month by Year
The average number of flights cancelled, delayed, diverted and on time per month, each year between 2003-2016.

```{r}
 
#Grouping by year
year <- delays_flights %>% group_by(Time.Year)

#Creating a data frame to store the summarized values of flights cancelled each year
delays_flights_year <- year %>% summarise(Statistics.Flights.Cancelled = mean(Statistics.Flights.Cancelled))

#Renaming the columns
colnames(delays_flights_year)[which(names(delays_flights_year) == "Time.Year")] <- "Year"
colnames(delays_flights_year)[which(names(delays_flights_year) == "Statistics.Flights.Cancelled")] <- "Cancelled"

#Summarizing by number of flights delayed each year
year_delayed <- year %>% summarise(Statistics.Flights.Delayed = mean(Statistics.Flights.Delayed))

#Adding the column to delays_flights_year
delays_flights_year$Delayed <-year_delayed$Statistics.Flights.Delayed

#Summarizing by number of flights delayed each year
year_delayed <- year %>% summarise(Statistics.Flights.Delayed = mean(Statistics.Flights.Delayed))
delays_flights_year$Delayed <-year_delayed$Statistics.Flights.Delayed

#Summarizing by number of flights diverted each year
year_diverted <- year %>% summarise(Statistics.Flights.Diverted = mean(Statistics.Flights.Diverted))
delays_flights_year$Diverted <-year_diverted$Statistics.Flights.Diverted

#Summarizing by number of flights on time each year
year_on_time <- year %>% summarise(Statistics.Flights.On.Time = mean(Statistics.Flights.On.Time))
delays_flights_year$On_Time <-year_on_time$Statistics.Flights.On.Time

#Creating a column to display percentage of flights that were on time each year, on average
delays_flights_year$Percent_On_Time <-100*delays_flights_year$On_Time/(delays_flights_year$On_Time+delays_flights_year$Diverted+delays_flights_year$Delayed+delays_flights_year$Cancelled)

#Display yearly averages for numbers of flights cancelled, delayed, diverted and on time
round(delays_flights_year) 
```

```{r}
#Identifying year with the greatest percent of flights on time, on overage
round(delays_flights_year[which.max(delays_flights_year$Percent_On_Time),])
```

```{r}
#Identifying year with the least percent of flights on time, on overage
round(delays_flights_year[which.min(delays_flights_year$Percent_On_Time),])
```

On average, flights were on time each month the most in <b>2012</b> and they were on time the least in <b>2007.</b>

# Delay Reason
## Average Number Flights Delayed per Month by Reason
The average percentage of flights delayed because of carrier, aircraft, national aviation system, security and weather delays between 2003-2016.
```{r}

#Average number of flights per month since 2003 delayed because of carriers
avg_flights_carriers <- sum(delays_reason$Statistics...of.Delays.Carrier)

#Average number of flights per month since 2003 delayed because of late arriving aircraft
avg_flights_aircraft <- sum(delays_reason$Statistics...of.Delays.Late.Aircraft)

#Average number of flights per month since 2003 delayed because of national aviation system
avg_flights_nas <- sum(delays_reason$Statistics...of.Delays.National.Aviation.System)

#Average number of flights per month since 2003 delayed because of security delays
avg_flights_security <- sum(delays_reason$Statistics...of.Delays.Security)

#Average number of flights per month since 2003 delayed because of weather
avg_flights_weather <- sum(delays_reason$Statistics...of.Delays.Weather)

#Creating a Pie Chart  
reason_values = c(avg_flights_carriers,avg_flights_aircraft,avg_flights_nas, avg_flights_security, avg_flights_weather)
reason_labels = c("Carrier","Late Aircraft","National Aviation System","Security","Weather")

#Calculating the percent each will be of the total
piepercent<- round(100*reason_values/sum(reason_values), 1)

# Plot the chart.
pie(reason_values, labels = piepercent, main = "Reasons for Flight Delay between 2003-2016",col = rainbow(length(reason_values)))
legend("topright", reason_labels, cex = 0.7,fill = rainbow(length(reason_values)))

```
The three biggest reasons for flight delays are 

- National Aviation System: <b>39.7%</b>
- Late Aircraft: <b>32.8%</b>
- Carrier: <b>23.9%</b>

## Delay Reason by Month of Year
```{r}
#Grouping by month
month <- delays_reason %>% group_by(Time.Month)

#Creating a data frame to store the summarized values of flight delayed because of carriers
delays_reason_month <- month %>% summarise(Statistics...of.Delays.Carrier = mean(Statistics...of.Delays.Carrier))

#Renaming the columns
colnames(delays_reason_month)[which(names(delays_reason_month) == "Time.Month")] <- "Month"
colnames(delays_reason_month)[which(names(delays_reason_month) == "Statistics...of.Delays.Carrier")] <- "Carrier"

#Summarizing by flights delayed because of late aircraft
month_aircraft <- month %>% summarise(Statistics...of.Delays.Late.Aircraft = mean(Statistics...of.Delays.Late.Aircraft))

#Adding the column to delays_flights_year
delays_reason_month$Aircraft <-month_aircraft$Statistics...of.Delays.Late.Aircraft

#Summarizing by flights delayed because of national aviation system
month_nas <- month %>% summarise(Statistics...of.Delays.National.Aviation.System = mean(Statistics...of.Delays.National.Aviation.System))
delays_reason_month$National_Aviation_System<-month_nas$Statistics...of.Delays.National.Aviation.System

#Summarizing by flights delayed because of security
month_security <- month %>% summarise(Statistics...of.Delays.Security = mean(Statistics...of.Delays.Security))
delays_reason_month$Security<-month_security$Statistics...of.Delays.Security

#Summarizing by flights delayed because of weather
month_weather <- month %>% summarise(Statistics...of.Delays.Weather = mean(Statistics...of.Delays.Weather))
delays_reason_month$Weather<-month_weather$Statistics...of.Delays.Weather

#Display yearly averages for numbers of flights cancelled, delayed, diverted and on time
round(delays_reason_month) 
```

```{r}
#Identifying month with the greatest number of delays due to each reasons for delay
round(delays_reason_month[which.max(delays_reason_month$National_Aviation_System),])
round(delays_reason_month[which.max(delays_reason_month$Aircraft),])
round(delays_reason_month[which.max(delays_reason_month$Carrier),])
round(delays_reason_month[which.max(delays_reason_month$Weather),])
round(delays_reason_month[which.max(delays_reason_month$Security),])
```

- <b>December</b> has the most delays happen due to the National Aviation System, Carrier and Security.
- <b>July</b> has the most delays due to Aircraft.
- <b>June</b> has the most delays due to Weather.

## Delay Reason by Year
```{r}
#Grouping by year
year <- delays_reason %>% group_by(Time.Year)

#Creating a data frame to store the summarized values of flight delayed because of carriers
delays_reason_year <- year %>% summarise(Statistics...of.Delays.Carrier = mean(Statistics...of.Delays.Carrier))

#Renaming the columns
colnames(delays_reason_year)[which(names(delays_reason_year) == "Time.Year")] <- "Year"
colnames(delays_reason_year)[which(names(delays_reason_year) == "Statistics...of.Delays.Carrier")] <- "Carrier"

#Summarizing by flights delayed because of late aircraft
year_aircraft <- year %>% summarise(Statistics...of.Delays.Late.Aircraft = mean(Statistics...of.Delays.Late.Aircraft))

#Adding the column to delays_flights_year
delays_reason_year$Aircraft <-year_aircraft$Statistics...of.Delays.Late.Aircraft

#Summarizing by flights delayed because of national aviation system
year_nas <- year %>% summarise(Statistics...of.Delays.National.Aviation.System = mean(Statistics...of.Delays.National.Aviation.System))
delays_reason_year$National_Aviation_System<-year_nas$Statistics...of.Delays.National.Aviation.System

#Summarizing by flights delayed because of security
year_security <- year %>% summarise(Statistics...of.Delays.Security = mean(Statistics...of.Delays.Security))
delays_reason_year$Security<-year_security$Statistics...of.Delays.Security

#Summarizing by flights delayed because of weather
year_weather <- year %>% summarise(Statistics...of.Delays.Weather = mean(Statistics...of.Delays.Weather))
delays_reason_year$Weather<-year_weather$Statistics...of.Delays.Weather

#Display yearly averages for numbers of flights cancelled, delayed, diverted and on time
round(delays_reason_year) 
```
```{r}
#Identifying year with the greatest number of delays due to each reasons for delay
round(delays_reason_year[which.max(delays_reason_month$National_Aviation_System),])
round(delays_reason_year[which.max(delays_reason_month$Aircraft),])
round(delays_reason_year[which.max(delays_reason_month$Carrier),])
round(delays_reason_year[which.max(delays_reason_month$Weather),])
round(delays_reason_year[which.max(delays_reason_month$Security),])
```
- <b>2014</b> has the most delays happen due to the National Aviation System, Carrier and Security.
- <b>2009</b> has the most delays due to Aircraft.
- <b>2008</b> has the most delays due to Weather.

# Delay Duration 
## Average Duration of Delay per Month by Reason
The average duration of delay because of carrier, aircraft, national aviation system, security and weather delays between 2003-2016.
```{r}

#Average duration of delay per month since 2003 delayed because of carriers
avg_flights_carriers <- mean(delays_duration$Statistics.Minutes.Delayed.Carrier)

#Average duration of delay per month since 2003 delayed because of late arriving aircraft
avg_flights_aircraft <- mean(delays_duration$Statistics.Minutes.Delayed.Late.Aircraft)

#Average duration of delay per month since 2003 delayed because of national aviation system
avg_flights_nas <- mean(delays_duration$Statistics.Minutes.Delayed.National.Aviation.System)

#Average duration of delay per month since 2003 delayed because of security delays
avg_flights_security <- mean(delays_duration$Statistics.Minutes.Delayed.Security)

#Average duration of delay per month since 2003 delayed because of weather
avg_flights_weather <- mean(delays_duration$Statistics.Minutes.Delayed.Weather)

#Creating a Bar Plot with values displayed on top of bars
duration_values = c(avg_flights_carriers,avg_flights_aircraft,avg_flights_nas, avg_flights_security, avg_flights_weather)
duration_values = round(duration_values)
duration_labels = c("Carrier","Late Aircraft","National Aviation System","Security","Weather")

## Make the duration values numbers (rather than factors)
duration_values <- as.numeric(as.character(duration_values))

## Find a range of y's that will leave sufficient space above the tallest bar
ylim <- c(0, 1.1*max(duration_values))

## Plot, and store x-coordinates of bars in xx
xx <- barplot(duration_values,main ="Average Duration of Delay per Month by Reason", col=rainbow(length(duration_values)), ylim = ylim)

## Add text at top of bars
text(x = xx, y = duration_values, label = duration_values, pos = 3, cex = 0.8, col = "red")

## Add x-axis labels 
axis(1, at=xx, labels=duration_labels, tick=FALSE, las=2, line=-0.5, cex.axis=0.5)
```
The longest delays were from three main sources:

- Late Aircraft resulted in the longest delays, averaging a total of 49,410 minutes each month.
- National Aviation System resulted in delays averaging a total of 45,077 minutes each month.
- Carriers resulted in delays averaging a total of 35,021 minutes each month.

## Delay Duration by Month of Year
```{r}
#Grouping by month
month <- delays_duration %>% group_by(Time.Month)

#Creating a data frame to store the summarized values of flight delayed because of carriers
delays_duration_month <- month %>% summarise(Statistics.Minutes.Delayed.Carrier = mean(Statistics.Minutes.Delayed.Carrier))

#Renaming the columns
colnames(delays_duration_month)[which(names(delays_duration_month) == "Time.Month")] <- "Month"
colnames(delays_duration_month)[which(names(delays_duration_month) == "Statistics.Minutes.Delayed.Carrier")] <- "Carrier"

#Summarizing by flights delayed because of late aircraft
month_aircraft <- month %>% summarise(Statistics.Minutes.Delayed.Late.Aircraft = mean(Statistics.Minutes.Delayed.Late.Aircraft))

#Adding the column to delays_flights_year
delays_duration_month$Aircraft <-month_aircraft$Statistics.Minutes.Delayed.Late.Aircraft

#Summarizing by flights delayed because of national aviation system
month_nas <- month %>% summarise(Statistics.Minutes.Delayed.National.Aviation.System=mean(Statistics.Minutes.Delayed.National.Aviation.System))
delays_duration_month$National_Aviation_System<-month_nas$Statistics.Minutes.Delayed.National.Aviation.System

#Summarizing by flights delayed because of security
month_security <- month %>% summarise(Statistics.Minutes.Delayed.Security = mean(Statistics.Minutes.Delayed.Security))
delays_duration_month$Security<-month_security$Statistics.Minutes.Delayed.Security

#Summarizing by flights delayed because of weather
month_weather <- month %>% summarise(Statistics.Minutes.Delayed.Weather = mean(Statistics.Minutes.Delayed.Weather))
delays_duration_month$Weather<-month_weather$Statistics.Minutes.Delayed.Weather

#Summarizing by total delay
month_total <- month %>% summarise(Statistics.Minutes.Delayed.Total = mean(Statistics.Minutes.Delayed.Total))
delays_duration_month$Total<-month_total$Statistics.Minutes.Delayed.Total

#Display yearly averages for numbers of flights cancelled, delayed, diverted and on time
round(delays_duration_month) 
```
```{r}
#Identifying month with the greatest duration of delays due to each reason
round(delays_duration_month[which.max(delays_duration_month$Total),])
round(delays_duration_month[which.max(delays_duration_month$National_Aviation_System),])
round(delays_duration_month[which.max(delays_duration_month$Aircraft),])
round(delays_duration_month[which.max(delays_duration_month$Carrier),])
round(delays_duration_month[which.max(delays_duration_month$Weather),])
round(delays_duration_month[which.max(delays_duration_month$Security),])
```
Overall, <b>July</b> has the longest delays.

- <b>June</b> has the longest delays due to the National Aviation System and Weather.
- <b>July</b> has the longest delays due to Aircraft and Carrier.
- <b>August</b> has the longest delays due to Security.

## Delay Duration by Year
```{r}
#Grouping by year
year <- delays_duration %>% group_by(Time.Year)

#Creating a data frame to store the summarized values of flight delayed because of carriers
delays_duration_year <- year %>% summarise(Statistics.Minutes.Delayed.Carrier = mean(Statistics.Minutes.Delayed.Carrier))

#Renaming the columns
colnames(delays_duration_year)[which(names(delays_duration_year) == "Time.Year")] <- "Year"
colnames(delays_duration_year)[which(names(delays_duration_year) == "Statistics.Minutes.Delayed.Carrier")] <- "Carrier"

#Summarizing by flights delayed because of late aircraft
year_aircraft <- year%>% summarise(Statistics.Minutes.Delayed.Late.Aircraft = mean(Statistics.Minutes.Delayed.Late.Aircraft))
delays_duration_year$Aircraft <-year_aircraft$Statistics.Minutes.Delayed.Late.Aircraft

#Summarizing by flights delayed because of national aviation system
year_nas <- year %>% summarise(Statistics.Minutes.Delayed.National.Aviation.System=mean(Statistics.Minutes.Delayed.National.Aviation.System))
delays_duration_year$National_Aviation_System<-year_nas$Statistics.Minutes.Delayed.National.Aviation.System

#Summarizing by flights delayed because of security
year_security <- year %>% summarise(Statistics.Minutes.Delayed.Security = mean(Statistics.Minutes.Delayed.Security))
delays_duration_year$Security<-year_security$Statistics.Minutes.Delayed.Security

#Summarizing by flights delayed because of weather
year_weather <- year %>% summarise(Statistics.Minutes.Delayed.Weather = mean(Statistics.Minutes.Delayed.Weather))
delays_duration_year$Weather<-year_weather$Statistics.Minutes.Delayed.Weather

#Summarizing by total delay
year_total <- year %>% summarise(Statistics.Minutes.Delayed.Total = mean(Statistics.Minutes.Delayed.Total))
delays_duration_year$Total<-year_total$Statistics.Minutes.Delayed.Total

#Display yearly averages for numbers of flights cancelled, delayed, diverted and on time
round(delays_duration_year) 
```
```{r}
#Identifying year with the greatest duration of delays due to each reason
round(delays_duration_year[which.max(delays_duration_year$Total),])
round(delays_duration_year[which.max(delays_duration_year$National_Aviation_System),])
round(delays_duration_year[which.max(delays_duration_year$Aircraft),])
round(delays_duration_year[which.max(delays_duration_year$Carrier),])
round(delays_duration_year[which.max(delays_duration_year$Weather),])
round(delays_duration_year[which.max(delays_duration_year$Security),])
```
Overall, <b>2007</b> has the longest delays.

- <b>2007</b> has the longest delays due to the National Aviation System, Aircraft, Carrier and Weather.
- <b>2006</b> has the longest delays due to Security.

# Summary of Data Analysis

On average, <b>77.8%</b> of flights were On Time each month and <b>20.2%</b> of flights were Delayed.

The three biggest reasons for flight delays are: 

- National Aviation System accounts for <b>39.7%</b> of all delays, averaging a total of 45,077 minutes each month.
- Late Aircraft accounts for <b>32.8%</b> of all delays, averaging a total of 49,410 minutes each month, which is the <b>longest</b> delay.
- Carrier accounts for <b>23.9%</b> of all delays, averaging a total of 35,021 minutes each month.

The most delays occur in the month of <b>December</b>, which can be accounted towards the National Aviation System, Carrier and Security.   

The longest delays occur in the month of <b>July</b>, which can be accounted towards the Aircraft and Carrier.

The most delays occur in the year <b>2007.</b>

- These delays can be accounted towards the National Aviation System, Aircraft, Carrier and Weather.
- This is also the year with the longest delays overall. 



